---
title: "P8105 Homework 2"
author: "Lina Marcinczyk"
date: "2026-10-01"
output: html_document
---

# Problem 1

```{r, message=FALSE}
library(tidyverse)
```

### Importing the NYC Transit dataset 
Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (case_match function may be useful).

```{r}
transit_df = 
  readr::read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
                  na = c("",".","NA")) |>
  janitor::clean_names()|> 
  dplyr::select(
    line, station_name, station_latitude, station_longitude, 
    route1:route11, entry, vending, entrance_type, ada
    ) |> 
  dplyr::mutate(
    entry = dplyr::case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE,
      .default = NA
    )) 
```

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

The cleaned subway dataset contains info about subway entrances, including the line, station name, routes served, whether entry is allowed, vending availability, entrance type, ADA compliance, and station coordinates. I standardized the column names, retained the relevant variables, and converted entry from YES/NO to TRUE/FALSE. The resulting dataset has 1,868 rows x 19 columns. The data are not fully tidy because routes served are spread across route1 through route11. These columns contain values of a single variable, route. A tidy structure would have one route column, with a separate row for each entrance–route combination. 

### Data questions
```{r}
transit_df |>
  distinct(line, station_name) |>
  nrow()
```
There are 465 distinct stations. 

```{r}
transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
84 of the stations are ADA compliant.

```{r}
transit_df |>
  filter(vending == "NO") |>
  summarise(
    proportion_allow_entry = mean(entry, na.rm = TRUE)
  )
```
The proportion of station entrances / exits without vending that 
allow entrance is 0.377.

### Reformat route number and name

```{r}
transit_long =
  transit_df |>
  mutate(across(route1:route11, as.character)) |>
  tidyr::pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  )
```
```{r}
transit_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```
There are 60 distinct stations that serve the A train.

```{r}
transit_long |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
17 of those 60 distinct stations are ADA compliant. 

# Problem 2

### Importing and cleaning Trash Wheels

```{r}
library(readxl)
library(dplyr)

mr_trash_df = read_excel(
  "data/Trash_Wheel_Collection_Data.xlsx", 
  sheet = "Mr. Trash Wheel",
  range = "A2:N710") |> 
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.integer(year),
    trash_wheel = "Mr. Trash Wheel"
  )

prof_trash_df = read_excel(
  "data/Trash_Wheel_Collection_Data.xlsx", 
  sheet = "Professor Trash Wheel",
  range = "A2:M135") |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    year = as.integer(year),
    trash_wheel = "Professor Trash Wheel"
  )

gwynnda_df = read_excel(
  "data/Trash_Wheel_Collection_Data.xlsx", 
  sheet = "Gwynns Falls Trash Wheel",
  range = "A2:L352") |> 
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |> 
  mutate(
    year = as.integer(year),
    trash_wheel = "Gwynnda"
    )

trash_wheel_df = bind_rows(
  mr_trash_df,
  prof_trash_df,
  gwynnda_df
)

trash_wheel_df

prof_total_weight = trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(total = sum(weight_tons, na.rm = TRUE)) |>
  pull(total)

gwynnda_june_2022_butts = trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    month == "June",
    year == 2022
  ) |>
  summarize(total = sum(cigarette_butts, na.rm = TRUE)) |>
  pull(total)
```

### Data summaries

The combined dataset contains `r nrow(trash_wheel_df)` observations, each representing a dumpster collected by Mr. Trash Wheel, Professor Trash Wheel, or Gwynnda. Key variables include the collection date, dumpster number, weight of trash in tons, volume in cubic yards, and counts of items such as plastic bottles and cigarette butts. The `trash_wheel` variable identifies which wheel collected each dumpster. Cleaning included standardizing variable names, removing rows without dumpster-specific data, and rounding Mr. Trash Wheel’s sports ball counts to integers. Across the available data, Professor Trash Wheel collected `r round(prof_total_weight, 2)` tons of trash, and Gwynnda collected `r gwynnda_june_2022_butts` cigarette butts in June 2022.

# Problem 3

Alzheimer's study

Discuss important steps in the import process and relevant features of the dataset. How many participants were recruited, and of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?

Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.

Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.

### Baseline Demographics

```{r}
baseline_df = 
  readr::read_csv(
    "data/MCI_baseline.csv",
                  skip = 1,
                  na = c("", ".", "NA")
    ) |> 
  janitor::clean_names() |> 
  dplyr::mutate(
    sex = factor(
      dplyr::case_match(
      sex,
      1 ~ "male",
      0 ~ "female"
      ))) |> 
   dplyr::mutate(
    apoe4 = factor(
      dplyr::case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
      ))) |> 
  filter(
    age_at_onset > current_age |
      is.na(age_at_onset)
  )
```


