---
title: "P8105 Homework 2"
author: "Lina Marcinczyk"
date: "2026-10-01"
output: html_document
---

# Problem 1

```{r, message=FALSE}
library(tidyverse)
```

### Importing the NYC Transit dataset 
Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (case_match function may be useful).

```{r}
transit_df = 
  readr::read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
                  na = c("",".","NA")) |>
  janitor::clean_names()|> 
  dplyr::select(
    line, station_name, station_latitude, station_longitude, 
    route1:route11, entry, vending, entrance_type, ada
    ) |> 
  dplyr::mutate(
    entry = dplyr::case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE,
      .default = NA
    )) 
```

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

The cleaned subway dataset contains info about subway entrances, including the line, station name, routes served, whether entry is allowed, vending availability, entrance type, ADA compliance, and station coordinates. I standardized the column names, retained the relevant variables, and converted entry from YES/NO to TRUE/FALSE. The resulting dataset has 1,868 rows x 19 columns. The data are not fully tidy because routes served are spread across route1 through route11. These columns contain values of a single variable, route. A tidy structure would have one route column, with a separate row for each entrance–route combination. 

### Data questions
```{r}
transit_df |>
  distinct(line, station_name) |>
  nrow()
```
There are 465 distinct stations. 

```{r}
transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
84 of the stations are ADA compliant.

```{r}
transit_df |>
  filter(vending == "NO") |>
  summarise(
    proportion_allow_entry = mean(entry, na.rm = TRUE)
  )
```
The proportion of station entrances / exits without vending that 
allow entrance is 0.377.

### Reformat route number and name

```{r}
transit_long =
  transit_df |>
  mutate(across(route1:route11, as.character)) |>
  tidyr::pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  )
```
```{r}
transit_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```
There are 60 distinct stations that serve the A train.

```{r}
transit_long |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
17 of those 60 distinct stations are ADA compliant. 

# Problem 2

### Importing and cleaning Trash Wheels

```{r}
library(readxl)
library(dplyr)

mr_trash_df = read_excel(
  "data/New_Trash_Wheel_Collection_Data.xlsx", 
  sheet = "Mr. Trash Wheel",
  range = "A2:N742") |> 
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.integer(year),
    trash_wheel = "Mr. Trash Wheel"
  )

prof_trash_df = read_excel(
  "data/New_Trash_Wheel_Collection_Data.xlsx", 
  sheet = "Professor Trash Wheel",
  range = "A2:M139") |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    plastic_bottles = as.numeric(na_if(plastic_bottles, "dive")),
    year = as.integer(year),
    trash_wheel = "Professor Trash Wheel"
  )

gwynnda_df = read_excel(
  "data/New_Trash_Wheel_Collection_Data.xlsx", 
  sheet = "Gwynnda the Good Wheel of the W",
  range = "A2:L394") |> 
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |> 
  mutate(
    year = as.integer(year),
    trash_wheel = "Gwynnda"
    )

trash_wheel_df = bind_rows(
  mr_trash_df,
  prof_trash_df,
  gwynnda_df
)

trash_wheel_df

prof_total_weight = trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(total = sum(weight_tons, na.rm = TRUE)) |>
  pull(total)

gwynnda_june_2022_butts = trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    month == "June",
    year == 2022
  ) |>
  summarize(total = sum(cigarette_butts, na.rm = TRUE)) |>
  pull(total)
```

### Data summaries

The combined dataset contains `r nrow(trash_wheel_df)` observations, each representing a dumpster collected by Mr. Trash Wheel, Professor Trash Wheel, or Gwynnda. Key variables include the collection date, dumpster number, weight of trash in tons, volume in cubic yards, and counts of items such as plastic bottles and cigarette butts. The `trash_wheel` variable identifies which wheel collected each dumpster. Cleaning included standardizing variable names, removing rows without dumpster-specific data, and rounding Mr. Trash Wheel’s sports ball counts to integers. Across the available data, Professor Trash Wheel collected `r round(prof_total_weight, 2)` tons of trash, and Gwynnda collected `r gwynnda_june_2022_butts` cigarette butts in June 2022.

# Problem 3

Alzheimer's study

### Baseline Demographics

```{r}
baseline_df = 
  readr::read_csv(
    "data/MCI_baseline.csv",
                  skip = 1,
                  na = c("", ".", "NA")
    ) |> 
  janitor::clean_names() |> 
  dplyr::mutate(
    sex = factor(
      dplyr::case_match(
      sex,
      1 ~ "male",
      0 ~ "female"
      ))) |> 
   dplyr::mutate(
    apoe4 = factor(
      dplyr::case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
      ))) |> 
  filter(
    age_at_onset > current_age |
      is.na(age_at_onset)
  ) 
```

During import, I had to skip row 1 in order for the clean_names function to work. The values marked as "." were recognized as NA, and sex and APOE4 needed to be converted to factors, as instructed. Lastly, 4 participants with MCI at baseline were excluded, while retaining the missing onset ages, as those participants qualified. 

There were 483 participants recruited, of whom `r nrow(baseline_df)` met the baseline inclusion criterion. Among eligible participants, `r sum(!is.na(baseline_df$age_at_onset))` developed MCI. The mean baseline age was `r round(mean(baseline_df$current_age, na.rm = TRUE), 2)` years. Among women with known APOE4 status, `r round(100 * mean(baseline_df$apoe4[baseline_df$sex %in% "female"] == "carrier", na.rm = TRUE), 1)`% were APOE4 carriers.

### Amyloid Measurements 

```{r}
messy_amyloid_df = 
  readr::read_csv(
    "data/mci_amyloid.csv",
                  skip = 1,
                  na = c("", ".", "NA", "Na")
    ) |> 
  janitor::clean_names() 

  amyloid_df= tidyr::pivot_longer(
      messy_amyloid_df,
      cols = c(baseline, time_2:time_8),
      names_to = "time",
      values_to = "amyloid_ratio"
    )
```

For import, I had to again skip row 1 for clean names, and also recognize "", "." as NA. Then I combined the amyloid ratio measurements into 1 column rather than 5 by using pivot_longer. 

The dataset contains 487 participants with measurements scheduled at baseline and years 2, 4, 6, and 8. After reshaping, it contains `r nrow(amyloid_df)` rows and `r ncol(amyloid_df)` columns: participant ID, visit time, and amyloid ratio. Each row represents one participant at a single scheduled visit. There are `r sum(is.na(amyloid_df$amyloid_ratio))` missing amyloid measurements, which were retained.

### Combining datasets

```{r}
amyloid_df = amyloid_df |> 
  dplyr::rename(id = study_id)

amyloid_ids <- amyloid_df |>
  dplyr::distinct(id) |>
  dplyr::mutate(has_amyloid = TRUE)

baseline_check <- baseline_df |>
  dplyr::left_join(amyloid_ids, by = "id")

baseline_check |>
  dplyr::filter(is.na(has_amyloid))

baseline_ids <- baseline_df |>
  dplyr::distinct(id) |>
  dplyr::mutate(has_baseline = TRUE)

amyloid_check <- amyloid_ids |>
  dplyr::left_join(baseline_ids, by = "id")

combined_df <- baseline_check |>
  dplyr::filter(has_amyloid == TRUE) |>
  dplyr::select(-has_amyloid) |>
  dplyr::left_join(amyloid_df, by = "id")
```

There were `r sum(is.na(baseline_check$has_amyloid))` eligible baseline participants without amyloid records and `r sum(is.na(amyloid_check$has_baseline))` amyloid participants without eligible baseline records. The combined dataset contains `r dplyr::n_distinct(combined_df$id)` participants, `r nrow(combined_df)` rows, and `r ncol(combined_df)` columns. Each row represents one participant at one scheduled visit, with baseline demographics and the amyloid measurement.

```{r}
readr::write_csv(combined_df, 
                 "data/mci_combined.csv")
```
